### 2022-03-16 model_discussion_report/model_discussion.tex
Finished method discussion report, updated lstm.py to include code to import all protein sequences

% NOTES: Could have a section talking explicitly about the concept of the "Memory Cell"
% COuld have a section at the end that discusses choices for model parameters in a specific example
% Could have section in beginning to go over the basics of recurrent neural networks

\documentclass[journal]{IEEEtran}
\usepackage{tikz}
\usetikzlibrary{decorations.pathreplacing,angles,quotes}

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
\title{LSTM Method Discussion Report}

% puts info about authors at bottom of first page
\author{Lawrence~Owusu,~Jordan~Sturtz,~and~Swetha~Chittam~% <-this % stops a space
  \thanks{The authors are graduate students at NCA\&T}% <-this % stops a space
}

% note the % following the last \IEEEmembership and also \thanks -
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
%
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!

% The paper headers
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
\markboth{CS851 - Deep Learning: Method Discussion Report}%
{Shell \MakeLowercase{\textit{et al.}}: CS851 - Deep Learning: Method Discussion Report}

% make the title area
\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
% \begin{abstract}
% The abstract goes here.
% \end{abstract}

% Note that keywords are not normally used for peerreview papers.
% \begin{IEEEkeywords}
% IEEE, IEEEtran, journal, \LaTeX, paper, template.
% \end{IEEEkeywords}

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi

%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Summary}
%
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{L}{ong} Short Term Memory (LSTM) is a powerful machine learning method
for learning tasks involving sequences with long-term dependencies. LSTM is a variant
of recurrent neural networks with a specialized architecture designed to solve the
vanishing gradient problem that prevents deep neural networks from learning long-term dependencies.
The original authors of LSTM, Hochreiter and Schmidhuber, designed the first variant of LSTM in 1997 that introduced
the idea of creating a unit in the neural network architecture that enforces constant error backflow [1]. To solve
the problem of conflicting weight signals, Hochreiter and Schmidhuber then added two additional features to their
model architecture: an input gate and an output gate [1]. Later, other authors would develop this architecture further
to add a "forget" gate, which is the variant of the LSTM architecture we explore in this report [2]. These gates use
a sigmoid layer together with pointwise multiplication to allow the network to filter signals.
\\

% FIXME: Not a good design pattern here to be manually entering reference numbers
% Should have an automatic method

\section{Recurrent Neural Networks}
Recurrent neural networks are neural networks that feature
feedback connections. These feedback connections typically occur when the output at time step $t$
is combined, usually concatenated, with the input for time step $t+1$ (See Fig. 1). LSTM
is a recurrent NN. Each memory block is recurrently connected to the next memory block,
which each memory block assigned to a specific timestep in the input sequence.

\begin{figure}[h]
  \centering
  \includegraphics[width=8cm]{RNN Illustration.png}
  \caption{Typical Simple RNN Structure}
\end{figure}

\section{The LSTM Architecture}
Characteristic of the LSTM architecture is a set of chained together cells called "memory blocks". The number
of memory blocks in the chain equals the length of the timesteps in our target dataset (Fig 2).
Each memory cell has three gating units (forget, input and output gates) which
conditionally regulate how information flows into and out of the memory block (Fig 3). Intuitively,
the "forget" gate can be viewed as a step where irrelevant information from the hidden state
is first "forgotten" before being passed to the input gate, which constructs the new cell state.
Before that cell state can output its value to the next memory cell, it is filtered again through
an output gate that decides what values to keep internal to the memory cell (i.e. the hidden state)
and what to output to the next memory cell or final output layer.

\begin{figure}[h]
\centering
\tikzstyle{place}=[rectangle,draw, thick, minimum width=30, minimum height=10]
\begin{tikzpicture}
    \node at (0,0) [place] (first) {Memory Cell};
    \node at (3,0) [place] (second) {Memory Cell};
    \node at (6,0) [place] (third) {Memory Cell};
    % \draw [->] (0,0) -- (first.west);
    \draw [->] (first.east) -- (second.west);
    \draw [->] (second.east) -- (third.west);
    % \draw [->] (third.east) -- (10,0);
    \draw[decoration={brace,mirror,raise=10pt},decorate](first.south west) -- node[below=16pt] {Number of cells = number of time steps} (third.south east);
\end{tikzpicture}
\caption{Illustration of recurrently chained memory cells}
\end{figure}

% \begin{figure}[h]
%   \centering
%   \includegraphics[width=8cm]{lstm_architecture.jpeg}
%   \caption{
%   The black dots indicate a sigmoidal activation function plus pointwise multiplication to produce a "gate" that
%   filters a signal by multiplying a value between 0 and 1 to decide the strength of a signal. The center of the
%   illustration shows a linear activation function which represents the "Constant Error Carosel" of the LSTM.
%   The feedback connection has a fixed weight of 1.0 on the previous timestep, with a linear activation function of
%   $f_{j}(x) = x, \forall x$, which Hochreiter and Schmidhuber refer to as the "Constant Error Carosel".
%   Sourced from Hochreiter and Schmidhuber, p. 1744}
% \end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=8cm]{jordan_lstm_architecture.png}
  \caption{LSTM architecture inside a memory cell}
\end{figure}

\subsection{The Forget Gate}
The forget gate decides what information to discard from the previous hidden state given
the new input data and previous hidden state. What inputs to ignore is learned in a sigmoid-activated
neural network that is then point multipled by the previous cell state, which effectively produces a "gate" that
filters the previous cell state. The results are then passed on to the next step. The output of the forget gate
can be expressed as $f_{t} = \sigma(W_{f}h_{t-1} + U_{f}X_{t} + b_{f})$,
where $W_{f}$, $U_{f}$ are weight matrices and $b_{f}$ is a bias vector [3, p. 555].

\subsection{The Input Gate}
The input gate in the LSTM is responsible for deciding which new
information to update the cell state given the new input data and the previous hidden state.
The new cell state update can be expressed with the following equations [3, p. 555]:

$\begin{array}{ll}
C_{t}   & = f_{t} \otimes C_{t-1} + i_{t}\bar{C} \\
i_{t}   & = \sigma(W_{t}h_{t-1} + U_{i}X_{t} + b_{i}) \\
\bar{C} & = tanh(W_{c}h_{t-1} + U_{c}X_{t} + b_{c})
\end{array}$

$W_{i}$, $U_{i}$ are weight matrices and $b_{i}$ is a bias vector. $C_{t}$ is the new cell state, and the above equation
represents this new state as a pointwise addition of the previous cell state after
running through the forget gate, $f_{t}C_{t-1}$, and the result of the current input gate, $i_{t}\bar{C}$.
The equation for $i_{t}$ is the scale factor applied to the candidate that could be added to the
internal state, represented by $\bar{C}$.

\subsection{The Output Gate}
Once the cell state has been updated, the output gate decides which components of the
cell state to keep "internal" to the memory cell.
The output gate step can be expressed by the following equations [3, p. 555]:

$\begin{array}{ll}
h_{t} & = o_{t}\otimes tanh(C_{t}) \\
o_{t} & = \sigma(W_{o}h_{t-1})
\end{array}$

The current cell state, $C_{t}$ is passed to a hyperbolic tangent function, which is then pointwise multiplied
by the output gate, $o_{t}$. The output function is again another sigmoid squashing function that
acts as a filter for the output to output the new hidden state.

\ifCLASSOPTIONcaptionsoff
  \newpage
\fi

\begin{thebibliography}{1}

\bibitem{IEEEhowto:kopka}
S.~Hochreiter and J.~Schmidhuber, "Long Short-Term Memory", \emph{Neural Computation}, vol. 9, no. 8, pp. 1735-1780, 1997.
\bibitem{IEEEhowto:kopka}
F.~Gers, J.~Schmidhuber, and F.~Cummins. "Learning to forget: Continual prediction with LSTM." \emph{Neural computation}, vol. 12, no. 10, 2451-2471, 2000.
\bibitem{IEEEhowto:kopka}
O~Calin. \emph{Deep learning architectures: A mathematical approach}. Switzerland: Springer, 2020. pp. 553-547.
% \bibitem{IEEEhowto:kopka}
% R.~Williams and D.~Zipser, "Gradient-based learning algorithms for recurrent networks and their computational complexity." in \emph{Backpropagation: Theory, architectures, and applications}, pp. 433-486.
% \bibitem{IEEEhowto:kopka}
% D.~Rumelhart, G.~Hinton, and R.~Williams, "Learning representations by back-propagating errors.", \emph{Nature}, vol. 323, no. 6088, 1735-1780, 1986

\end{thebibliography}

\end{document}

### 2022-03-16 model_discussion_report/lstm.py

@@ -1,0 +1,85 @@
+
+import numpy
+from keras.models import Sequential
+from keras.layers import Dense, LSTM, Conv1D
+from keras.utils import np_utils
+
+# fix random seed for reproducibility
+numpy.random.seed(7)
+
+# import the data
+def get_names_to_sequences(filename):
+    with open(filename, "r") as input_file:
+        sequences = [seq.split("\n") for seq in input_file.read().split(">") if seq]
+        return {name: "".join(parts).strip() for name, *parts in sequences}
+
+names_to_sequences = get_names_to_sequences("alemtuzumab_sequences.txt")
+
+# create mapping of characters to integers and the reverse for all characters in all sequences
+all_chars = set("".join(list(names_to_sequences.values())))
+char_to_int = {c: i for i, c in enumerate(all_chars)}
+int_to_char = {v: k for k, v in char_to_int.items()}
+
+# Model parameters
+kmer_length = 10
+epoch_length = 5
+num_lstm_layers = 256
+first_conv_filters = 128
+first_conv_kernel_size = 5
+first_dense_layer = 128
+second_dense_layer = 64
+cost_func = "categorical_crossentropy"
+optimizer = "adam"
+output_activation_func = "softmax"
+hidden_layer_activation_func = "relu"
+
+for test_name in names_to_sequences:
+    test_seq = names_to_sequences[test_name]
+    training_seqs = [names_to_sequences[name] for name in names_to_sequences if name != test_name]
+
+    train_as_lstX = []
+    train_as_lstY = []
+    test_as_lstX = []
+    test_as_lstY = []
+
+    # Iterate over each sequence, starting at kmer_length to ignore first kmer_length characters
+    for seq in training_seqs:
+        for index in range(kmer_length, len(seq)):
+            seq_in = seq[index - kmer_length:index]
+            seq_out = seq[index]
+            train_as_lstX.append([char_to_int[char] for char in seq_in])
+            train_as_lstY.append(char_to_int[seq_out])
+
+    for index in range(kmer_length, len(test_seq)):
+        seq_in = test_seq[index - kmer_length:index]
+        seq_out = test_seq[index]
+        test_as_lstX.append([char_to_int[char] for char in seq_in])
+        test_as_lstY.append(char_to_int[seq_out])
+
+    # reshape X to be [samples, time steps, features]
[…]

### 2022-03-17 model_discussion_report/lstm.py
Updated code to use test / validation data and to report training / validation error to check for overtraining

@@ -1,8 +1,10 @@
-
+import json
 import numpy
 from keras.models import Sequential
 from keras.layers import Dense, LSTM, Conv1D
 from keras.utils import np_utils
+from matplotlib import pyplot as plt
+from matplotlib.ticker import MaxNLocator
 
 # fix random seed for reproducibility
 numpy.random.seed(7)
@@ -11,75 +13,120 @@ numpy.random.seed(7)
 def get_names_to_sequences(filename):
     with open(filename, "r") as input_file:
         sequences = [seq.split("\n") for seq in input_file.read().split(">") if seq]
-        return {name: "".join(parts).strip() for name, *parts in sequences}
+        return {name.replace(":", "").replace("|", ""): "".join(parts).strip() for name, *parts in sequences}
 
-names_to_sequences = get_names_to_sequences("alemtuzumab_sequences.txt")
+# Save what we want from the results
+def save_result(filename, testname, test_accuracy, history, epoch_length):
+    epoch_axis = list(range(1, epoch_length + 1))
 
-# create mapping of characters to integers and the reverse for all characters in all sequences
-all_chars = set("".join(list(names_to_sequences.values())))
-char_to_int = {c: i for i, c in enumerate(all_chars)}
-int_to_char = {v: k for k, v in char_to_int.items()}
+    # Note that even in the OO-style, we use `.pyplot.figure` to create the Figure.
+    fig, ax = plt.subplots()
+    ax.plot(epoch_axis, history["accuracy"], label='Training')
+    ax.plot(epoch_axis, history["val_accuracy"], label='Validation')
+    ax.set_xlabel('Epochs')  # Add an x-label to the axes.
+    ax.set_ylabel('Accuracy')  # Add a y-label to the axes.
+    ax.set_title(f'Accuracy on {testname}')  # Add a title to the axes.
+    ax.xaxis.set_major_locator(MaxNLocator(integer=True))
 
-# Model parameters
-kmer_length = 10
-epoch_length = 5
-num_lstm_layers = 256
-first_conv_filters = 128
-first_conv_kernel_size = 5
-first_dense_layer = 128
-second_dense_layer = 64
-cost_func = "categorical_crossentropy"
-optimizer = "adam"
-output_activation_func = "softmax"
-hidden_layer_activation_func = "relu"
+    ax.annotate(f'Final Test Accuracy: {test_accuracy}',
+                xy = (1.0, -0.2),
+                xycoords='axes fraction',
+                ha='right',
+                va="center",
+                fontsize=10)
[…]

### 2022-04-05 lstm.py
Removed approaches logic, added path for results directory to write contents, fixed save_results

@@ -1,6 +1,10 @@
 import json
 import numpy as np
 import random
+import os
+import time
+from datetime import datetime
+
 from keras.models import Sequential
 from keras.layers import Dense, LSTM, Conv1D
 from keras.utils import np_utils
@@ -8,9 +12,6 @@ from matplotlib import pyplot as plt
 from matplotlib.ticker import MaxNLocator
 from tensorflow.keras.callbacks import EarlyStopping
 from tensorflow.keras.callbacks import Callback
-from datetime import datetime
-import time
-from enum import Enum
 
 # Use monospace to make formatting cleaner
 plt.rcParams.update({'font.family':'monospace'})
@@ -22,9 +23,12 @@ np.random.seed(7)
 # Globals here
 # =============================================================================
 
+# Path to save results
+RESULTS_PATH = os.path.join(os.getcwd(), "results")
+
 # Model parameters
 KMER_LENGTH = 20
-MAX_EPOCH_LENGTH = 500
+MAX_EPOCH_LENGTH = 5
 NUM_LSTM_LAYERS = 512
 FIRST_CONV_FILTERS = 128
 FIRST_CONV_KERNEL_SIZE = 5
@@ -63,21 +67,16 @@ all_chars = set("".join(list(names_to_sequences.values())))
 char_to_int = {c: i for i, c in enumerate(all_chars)}
 num_classes = len(char_to_int)
 
-# Used in formatting table of parameters
-# pad_name = max(len(item) for item in parameters.keys())
-# pad_value= max(len(str(item)) for item in parameters.values())
-
 # =============================================================================
 # FUNCTIONS START HERE
 # =============================================================================
 
 # =============================================================================
 # Save what we want from the results
-def save_result(run_number, approach, test_accuracy, history, model, secs_to_train):
+def save_result(test_accuracy, history, model, secs_to_train):
     now = datetime.now().strftime("%H-%M-%S")
     epoch_length = len(history["epochs_to_patience"])
     epoch_axis = list(range(1, epoch_length + 1))
-    pretty_approach_name = " ".join(item.capitalize() for item in approach.split("_"))
 
     # Note that even in the OO-style, we use `.pyplot.figure` to create the Figure.
     fig, ax = plt.subplots()
@@ -85,7 +84,7 @@ def save_result(run_number, approach, test_accuracy, history, model, secs_to_tra
[…]

### 2022-04-06 paper_discussion_report/paper_discussion.tex
Added in Swetha's content for the model architecture / parameters / results / etc

\documentclass[journal]{IEEEtran}
\usepackage{tikz}
\usetikzlibrary{decorations.pathreplacing,angles,quotes}

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
\title{Paper Discussion Report}

% puts info about authors at bottom of first page
\author{Lawrence~Owusu,~Jordan~Sturtz,~and~Swetha~Chittam~% <-this % stops a space
  \thanks{The authors are graduate students at NCA\&T}% <-this % stops a space
}

% note the % following the last \IEEEmembership and also \thanks -
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
%
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!

% The paper headers
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
\markboth{CS851 - Deep Learning: Paper Discussion Report}%
{Shell \MakeLowercase{\textit{et al.}}: CS851 - Deep Learning: Method Discussion Report}

% make the title area
\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
% \begin{abstract}
% The abstract goes here.
% \end{abstract}

% Note that keywords are not normally used for peerreview papers.
% \begin{IEEEkeywords}
% IEEE, IEEEtran, journal, \LaTeX, paper, template.
% \end{IEEEkeywords}

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi

%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Summary}

%
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{A}{} paper authored by Gunasekaran, et al. (2021) uses deep learning methods 
to perform classification of viruses from their DNA sequences. Since DNA is composed of strings
of nucleotides, the problem amounts to classifying viruses according to samples of nucleotide
strings. The authors collect data from the public nucleotide sequence database,
The National Centre for Biotechnology Information (NCBI) https://www.ncbi.nlm.nih.gov.
They then encode this data using label encoding and kmer encoding. For each encoding type,
they run three different deep learning models: CNN, CNN-LSTM, and CNN-Bidirectional-LSTM.
The architectures of all three models start with embedding layers, then convolutional layers,
then max pooling layers, and then from there diverge to either add LSTM layers or 
bidrectional LSTM layers before finishing with dense layers and a final output layer.
The authors compare all six combinations of the two encoding methods with the three model types
using several performance metrics.

\section{Problem Statement}
All DNA and RNA is composed of a string of nucleotides. A nucleotide refers to one of
four compounds for DNA (adenine, cytocine, guanine, thymine) or four compounds for RNA
(adenine, cytocine, guanine, uracil). For double-helix DNA or RNA, each nucleotide
bonds with one and only one other nucleotide, forming what is called a base pair (Fig 1). Since
these base pairs are fixed, then, a DNA or RNA sequence can be identified solely by one
side of the double helix. Thus, every DNA virus can be identified by a single string of
characters drawn from the set $\{A, C, G, T\}$ and every RNA virus can be identified by a string
of characters drawn from the set $\{A, C, G, U\}$. The task, then, is to build highly accurate
models to classify a virus from its DNA or RNA sample.

\begin{figure}
  \centering
  \includegraphics[width=8cm]{figures/dna.png}
  \caption{Single or double-stranded DNA/RNA, borrowed from Gunasekaran, H., et al. (2020)}
\end{figure}

\section{Related Work}

\subsection{Current Results on Proposed Problem}

\section{Data Collection}
The authors obtain complete genomic sequences from the 
National Centre for Biotechnology Information (NCBI) https://www.ncbi.nlm.nih.gov/.
Sequence length ranges from 8 to 37971 nucleoids. They collected genomic sequences
for six virus classes: COVID, MERS, SARS, Dengue, Hepatitus, and Influenza (Fig 2).
Because the population of these viruses were unbalanced--for instance, there were 37272
samples of COVID and only 1418 samples of MERS--the authors opted to use
Synthetic Minority Oversampling Technique (SMOTE) to get a more even distribution of
all six classes in their dataset.

\begin{figure}
  \centering
  \includegraphics[width=8cm]{figures/data_collection.png}
  \caption{Samples of virus classes retrieved from NCBI, borrowed from Gunasekaran, H., et al. (2020)}
\end{figure}

\section{Data Preprocessing}
The authors encoded the data in two different formats for comparative analysis.
In the first approach, they use label encoding, which replaces each nucleoid by a unique index value,
preserving positional information (Fig 3).
In the second approach, they used kmer encoding, which generates all kmers from a sequence 
and forms an English-like sentence onto which natural language processing techniques
can be applied (Fig 4).

\begin{figure}
  \centering
  \includegraphics[width=8cm]{figures/label_encoding.png}
  \caption{Label encoding example, borrowed from Gunasekaran, H., et al. (2020)}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=8cm]{figures/kmer_encoding.png}
  \caption{Kmer encoding example, where k=6}
\end{figure}

Once encoded, in both cases the input data is one-hot encoded and then fed into
the first layer of the models, which is an embedding layer.

\section{Proposed Models}
In this paper, the authors use three different classification models to perform multi-classification 
of the DNA sequence: CNN, CNN-LSTM and CNN-bidirectional LSTM.
All three models share the same initial layers: an embedding layer, convolutional layers, and
max pooling layers. The raw data is first either label encoded or kmer encoded and then 
one-hot encoded to be fed into the embedding layer. The results of the embedding layer
are then fed into the convolutional layers, each of which has a corresponding max pooling layer.
From here, the three model types diverge. For the CNN model, the results are fed into
the final dense layers and output layers. The LSTM and bidrectional LSTM models differ only
in that they add LSTM or bidirectional LSTM respectively before the final dense layers.

\subsection{CNN}
CNN is a popular deep learning technique that is widely used for feature extraction 
and to solve any classification problems. 
CNN can be used for any kind of datasets like text, image, or video datasets. 
The 2D CNN is often used for feature extraction from image datasets and 
3D CNN is used for feature extraction from video datasets. 
The 1D CNN auto extracts the features from the input dataset and can therefore be used 
for text classification. Here in this paper, authors are using 1D CNN for feature extraction 
of the DNA sequence data. The proposed CNN model architecture has two 1D convolutional layers,
each followed by a max pooling layer to reduce the feature map dimensions from the previous layers.
Next, a flatten layer is used to convert feature maps to a single column vector followed by multiple dense layers. 
The output is passed to the final dense layer with softmax activation function to perform multi-classification.

Table I shows the hyper parameters for the proposed CNN model
\begin{table}
  \caption{\label{tab:table-name}Hyper parameters of CNN model}
  \centering
  \resizebox{\columnwidth}{!}{%
    \begin{tabular}{ || l l l l l l l ||}
    \hline
    Layers           & Units & Filters & Kernel & Act. Func & Output Shape & Params   \\ 
    \hline
    \hline
    Embedding        & 8     &         &        &           & (None, 1000, 8)   & 128   \\  
    Conv 1D          &       & 128     & 2*2    & ReLu      & (None, 1000, 128) & 3200  \\
    Max Pool         &       &         & 2*2    &           & (None, 500, 128)  & 0     \\
    Conv 1D          &       & 64      & 2*2    & ReLu      & (None, 500, 64)   & 24640 \\
    Max Pool         &       &         & 2*2    &           & (None, 250, 64)   & 0     \\
    Flatten          &       &         &        &           & (None, 16)        & 0     \\
    Dense            & 128   &         &        &           & (None, 128)       & 2176  \\
    Dense            & 64    &         &        &           & (None, 64)        & 8256  \\
    Dense            & 6     &         &        & Softmax   & (None, 6)         & 390   \\
    \hline
    \end{tabular}%
  }
\end{table}

\subsection{CNN-LSTM}
Long-short term memory (LSTM) is a recurrent neural network that can learn long term dependencies in a sequence and 
hence can be used for sequence classification or prediction. The LSTM model consists of three gates: forget gate, input gate and output gate (Fig 5).
For the hybrid CNN-LSTM model, the authors add an LSTM Layer with 100 LSTM memory units
between the final convolutional layer and the final dense layers.

\begin{figure}
  \centering
  \includegraphics[width=8cm]{figures/lstm_cell.png}
  \caption{LSTM Cell Architecture. x is a multiplicative gate, + is additive, borrowed from Gunasekaran, H., et al. (2020)}
\end{figure}

% I think this explanation could be better. Also might not need to be included
% The current state is a function of both the previous hidden state and the
% current input: $h(t) = f(h(t-1), x(t))$.
% The forget gate is responsible for removing any information from the cell state. 
% When the information becomes invalid, it outputs zero. 
% The input gate is responsible for adding relevant information to the cell state, 
% this gate uses two activation functions sigmoid and tanh. 
% The output gate is responsible to output the current cell state h(t) by employing sigmoid and tanh 
% activation functions.

\subsection{CNN-Bidirectional-LSTM}
The bi-directional LSTM has two RNN’s: One to learn dependencies in forward direction and the other to 
learn dependencies in backward direction (Fig 6). For the hybrid CNN-LSTM-Bidirectional model, the authors add the bidirectional LSTM layers
after the final convolutional layers and before the final dense layers.

\begin{figure}
  \centering
  \includegraphics[width=8cm]{figures/bi_lstm_cell.png}
  \caption{Bidirectional LSTM Model Architecture, borrowed from Gunasekaran, et al. (2020)}
\end{figure}

\section{Results and Discussion}
The DNA sequence dataset used in this paper consists of 66,153 inputs which is divided into training, 
validation and testing sets with a ratio of 70\%, 10\% and 20\% respectively.
The training set consists of 46307, and the validation set consists of 6615, and the testing set consists of 13231 samples. 
The maximum sequence length is 2000, and the vocabulary size is 8972. 
The proposed models CNN, CNN-LSTM, CNN-Bi-LSTM models are tested by varying different hyperparameters listed in Table I. 
The authors use the grid-search cross-validation technique to find the best parameters of the model.
The best parameters are listed in Table II.

\begin{table}
  \caption{\label{tab:table2}Best Hyperparameters}
  \centering
  % \resizebox{\columnwidth}{!}{%
    \begin{tabular}{ || l l ||}
    \hline
    Parameters & Values \\ 
    \hline
    \hline
    Size of Filter       & 2*2 \\  
    Training Batch Size  & 100 \\
    Training Epochs      & 10  \\
    Embedding Dimensions & 32  \\
    K-mer Size           & 6   \\
    \hline
    \end{tabular}%
  % }
\end{table}

The classification models are evaluated using different classification metrics like accuracy, precision, recall, F1 score, sensitivity, 
and specificity by obtaining the confusion matrix for both k-mer and label encoding techniques.
The confusion matrix holds the values of true positives (TP), true negatives (TN), false positives (FP), and false negatives(FN). 
Based on these values, classification metrics are calculated and compared.
Table III shows the formulas for calculating these classification metrics. 
\begin{table}
  \caption{\label{tab:table3}Formulas to Calculate Performance Metrics}
  \centering
  % \resizebox{\columnwidth}{!}{%
    \begin{tabular}{ || l l ||}
    \hline
    Metric & Formula \\ 
    \hline
    \hline
    Accuracy    & $\frac{TP + TN}{TP + TN + FP + FN}$\\  
    Specificity & $\frac{TN}{TN + FP}$ \\
    Sensitivity & $\frac{TP}{TP + FN}$ \\
    Precision   & $\frac{TP}{TP + FP}$ \\
    \hline
    \end{tabular}%
  % }
\end{table}

\subsection{Model Comparison}
Figure 7 shows the accuracy comparison for all three models for training and testing data using label and k-mer encoding. 
The testing accuracies for label encoding are less when compared to its training accuracy. 
For k-mer encoding testing accuracies are more significant than training. 
Therefore, the encoding technique plays an important role for achieving high accuracy.

\begin{figure}
  \centering
  \includegraphics[width=8cm]{figures/accuracy.png}
  \caption{Bidirectional LSTM Model Architecture, borrowed from Gunasekaran, et al. (2020)}
\end{figure}

Figure 8 shows the training and validation accuracies for all three models and both encodings. 
From Figure 8, we can observe that the accuracy of the model remains same after 10 epochs for all models 
except CNN-LSTM using k-mer encoding. This model has unstable accuracy as each epoch increases.
CNN and CNN-Bi-LSTM have high accuracies of 93.16\%, 93.13\% respectively, when compared to LSTM.

% FIXME: I wanted to use figure* environment to span both columns, but the damn thing is broke
% It won't stay positioned "here". The documentation discusses figure* but doesn't give much insight
% into how to fix. So for now, let this span a single column, which makes it look super fooking tiny
\begin{figure}
\centering
  \includegraphics[width=8cm]{figures/train_vrs_valid_accuracies.png}
  \caption{Training / Validation accuracies for all six models, borrowed from Gunasekaran, et al. (2020)}
\end{figure}

\subsection{Model Performance Against State of the Art Methods}

The authors compare their accuracy results to other state-of-the-art approaches to 
virus classification (Fig 9) Their results show that all three of their models show higher accuracy than
that achieved by Nguyen, et al. (2016), Do, et al. (2020), and Zhang, et al. (2020).

\begin{figure}
  \centering
  \includegraphics[width=8cm]{figures/accuracy_compared_to_state_of_art.png}
  \caption{Gunasekaran, H. et al, compared against state of the art methods, borrowed from Gunasekaran, et al. (2020)}
\end{figure}

% 3.1 Model comparison :
% The proposed models are compared with different techniques from [1,2,3] to prove the robustness of the models. 
% The achieved accuracy is obtained in classifying six classes. 
% It is observed that all three models have achieved higher accuracies when compared to the state-of-the-art methods.

% Figure 6: Comparison of the models with the state of the art methods.

% Table 4 : Model performance concerning different classes using label and k-mer encoding techniques.

% Table 4 shows the model performance  concerning six different classes using label and k-mer encoding techniques. Based on metrics in Table 4, we can conclude that 
% ⦁	If class samples are high, CNN with label encoding has high precision.
% ⦁	If class samples are low, CNN with k-mer encoding has high precision.
% ⦁	If recall is preferred, k-mer encoding has higher recall for all classes irrespective of the class samples.  
% ⦁	High sensitivity rate of 99.95% is obtained for class “a” using CNN+Bi-LSTM with label encoding.
% ⦁	To obtain a high recall and sensitivity for class with more samples, Bi-LSTM with label encoding will be a good choice.
% ⦁	CNN with label encoding offers high specificity irrespective of the class size.
% Conclusion :
% This paper compared three models with label and k-mer encoding techniques. CNN with label encoding outperforms other two models but their testing accuracies are low. K-mer encoding has achieved high validation and testing accuracy. We can conclude that, the best model can be chose not only based on the accuracy of the model but also other metrics such as sensitivity, specificity and precision must be considered.
% References :
% ⦁	M. F. Aslan, M. F. Unlersen, K. Sabanci, and A. Durdu, “CNNbased transfer learning-BiLSTM network: a novel approach for COVID-19 infection detection,” Applied Soft Computing, vol. 98, article 106912, 2021.

% ⦁	X. Zhang, B. Beinke, B. Al Kindhi, and M. Wiering, “Comparing machine learning algorithms with or without feature extraction for DNA classification,” 2020, http://arxiv.org/abs/2011.00485.

% ⦁	S. W. I. Maalik and S. K. W. Ananta, “Comparation analysis of ensemble technique with boosting (Xgboost) and bagging (Randomforest) for classify splice junction DNA sequence Computational and Mathematical Methods in Medicine 11 category,” Jurnal Penelitian Pos dan Informatika, vol. 9, no. 1,pp. 27–36, 2019.
\section{Conclusion}

\begin{thebibliography}{1}

\bibitem{IEEEhowto:kopka}
H.~Gunasekaran, K.~Ramalakshmi, A.~Rex~Macedo~Arokiaraj, S.~Deepa~Kanmani, C.~Venkatesan, and C.~Suresh~Gnana~Dhas.
  "Analysis of DNA Sequence Classification Using CNN and Hybrid Models." 
  \emph{Computational and mathematical methods in medicine}, 2021.
\bibitem{IEEEhowto:kopka}
N.~G.~Nguyen,~V.~A.~Tran,~D.~L.~Ngo et al., “DNA sequence classification by convolutional neural network,” \emph{Journal of Bio-
medical Science and Engineering}, vol. 9, no. 5, pp. 280–286, 2016.
\bibitem{IEEEhowto:kopka}
D.~T.~Do~and~N.~Q.~K.~Le, “Using extreme gradient boosting to identify origin of replication in Saccharomyces cerevisiae via
hybrid features,” \emph{Genomics}, vol. 112, no. 3, pp. 2445–2451,2020
\bibitem{IEEEhowto:kopka}
X.~Zhang,~B.~Beinke,~B.~Al~Kindhi,~and~M.~Wiering, “Comparing machine learning algorithms with or without feature extraction for DNA classification"
2020, http://arxiv.org/abs/2011.00485
\end{thebibliography}

\end{document}

### 2022-04-07 paper_discussion_report/paper_discussion.tex
Incorporated Lawrence's contribution, reorganized everything

@@
   % and "HIS" in caps to complete the first word.
-\IEEEPARstart{A}{} paper authored by Gunasekaran, et al. (2021) uses deep learning methods 
-to perform classification of viruses from their DNA sequences. Since DNA is composed of strings
+  \IEEEPARstart{A}{} paper authored by Gunasekaran, et al. uses deep learning methods 
+  to classify viruses by their DNA sequences[1]. Since DNA is composed of strings
   of nucleotides, the problem amounts to classifying viruses according to samples of nucleotide
@@
   then max pooling layers, and then from there diverge to either add LSTM layers or 
-bidrectional LSTM layers before finishing with dense layers and a final output layer.
+  bidirectional LSTM layers before finishing with dense layers and a final output layer.
   The authors compare all six combinations of the two encoding methods with the three model types
-using several performance metrics.
+  using several performance metrics. They also compare their results to current state-of-the-art
+  approaches in the literature.
 
@@
     \includegraphics[width=8cm]{figures/dna.png}
-  \caption{Single or double-stranded DNA/RNA, borrowed from Gunasekaran, H., et al. (2020)}
+    \caption{Single or double-stranded DNA/RNA, borrowed from Gunasekaran, H., et al.[1]}
   \end{figure}
@@
 \section{Related Work}
+  DNA classification and protein structure-sequence prediction remain some of the challenges in the field of computational biology. 
+  Determination of accurate protein structure and DNA sequence are important in developing deep understanding of the functions of protein,
+  disease diagnosis, prevention of epidemics and assisting in drug design [2]. 
+  \subsection{Non-Machine Learning Methods}
 
-\subsection{Current Results on Proposed Problem}
+    Although experimental methods of protein structure determination, including Cryogenic Electron Microscopy, 
+    X-ray Crystallography, Nuclear Magnetic Resonance Spectroscopy and others have seen continuous advancement, 
+    such techniques are expensive, involve cumbersome and time-consuming procedures and require high level expertise. 
+    Similarly, DNA-based approaches such as DNA sequencing, DNA-DNA hybridization and DNA fingerprinting for the identification 
+    and classification of species of bacteria and viruses have some challenges. For example, these DNA sequencing technologies 
+    produce read length as short as 35-40 nucleotides, thus posing challenges for genome annotation and assembly [3].
+    To make the process less cumbersome, deep learning algorithms as well as hybrid deep learning models have been proposed 
+    for DNA and protein sequence classification [4].
+
+  \subsection{Machine Learning Methods}
+    The use of CNN and LSTM models in protein sequence prediction, classification of miRNA and DNA sequence classification is 
+    highly researched in computational biology. Abdulkadir and Baha (2021) applied a hybrid deep learning model based on 
+    CNN and LSTM to classify miRNA with accuracy, sensitivity, specificity and F1 scores of 0.943, 0.935, 0.948 and 0.925 respectively, 
+    suggesting that the hybrid CNN and LSTM networks can be employed to achieve better performance for pre-miRNA classification [5].
+    Similarly, Chinju and his colleagues (2022) developed a hybrid deep learning CNN-LSTM model that uses the 
+    convolutional neural network architecture and long short-term memory network to classify the protein sequences belonging to 
+    the five-member polo like kinase family with an accuracy score of 97.6\% [6]. 
+    LSTM networks have been used tremendously in computational biology especially in protein structure prediction. 
+    For example, Soren and Winther (2019) reported 67.4\% performance accuracy on protein structure prediction problems [7].
+
+\subsection{Model Performance Against State of the Art Methods}
+
+  The authors compare their accuracy results to other state-of-the-art approaches to 
+  virus classification (Fig 2). Their results show that all three of their models show higher accuracy than
+  that achieved by Nguyen, et al.[8], Do, et al.[9], and Zhang, et al.[10].
+
+  \begin{figure}
+    \centering
+    \includegraphics[width=8cm]{figures/accuracy_compared_to_state_of_art.png}
+    \caption{Gunasekaran, H. et al, compared against state of the art methods, borrowed from Gunasekaran, et al. (2020)}
+  \end{figure}
 
@@
   Sequence length ranges from 8 to 37971 nucleoids. They collected genomic sequences
-for six virus classes: COVID, MERS, SARS, Dengue, Hepatitus, and Influenza (Fig 2).
+  for six virus classes: COVID, MERS, SARS, Dengue, Hepatitus, and Influenza (Fig 3).
   Because the population of these viruses were unbalanced--for instance, there were 37272
@@
   all six classes in their dataset.
+  The DNA sequence dataset used in this paper consists of 66,153 inputs which is divided into training, 
+  validation and testing sets with a ratio of 70\%, 10\% and 20\% respectively.
+  The training set consists of 46307, and the validation set consists of 6615, and the testing set consists of 13231 samples. 
+  The maximum sequence length is 2000, and the vocabulary size is 8972.
 
@@
   In the first approach, they use label encoding, which replaces each nucleoid by a unique index value,
-preserving positional information (Fig 3).
+  preserving positional information (Fig 4).
   In the second approach, they used kmer encoding, which generates all kmers from a sequence 
   and forms an English-like sentence onto which natural language processing techniques
-can be applied (Fig 4).
+  can be applied (Fig 5).
 
@@
     \includegraphics[width=8cm]{figures/kmer_encoding.png}
-  \caption{Kmer encoding example, where k=6}
+    \caption{Kmer encoding example, where k=6, borrowed from Gunasekaran, H., et al. (2020)}
   \end{figure}
@@
   From here, the three model types diverge. For the CNN model, the results are fed into
-the final dense layers and output layers. The LSTM and bidrectional LSTM models differ only
+  the final dense layers and output layers. The LSTM and bidirectional LSTM models differ only
   in that they add LSTM or bidirectional LSTM respectively before the final dense layers.
@@
 
-Table I shows the hyper parameters for the proposed CNN model
+    Table I shows the hyper parameters for the proposed CNN model.
     \begin{table}
@@
     Long-short term memory (LSTM) is a recurrent neural network that can learn long term dependencies in a sequence and 
-hence can be used for sequence classification or prediction. The LSTM model consists of three gates: forget gate, input gate and output gate (Fig 5).
+    hence can be used for sequence classification or prediction. The LSTM model consists of three gates: forget gate, input gate and output gate (Fig 6).
     For the hybrid CNN-LSTM model, the authors add an LSTM Layer with 100 LSTM memory units
@@
     The bi-directional LSTM has two RNN’s: One to learn dependencies in forward direction and the other to 
-learn dependencies in backward direction (Fig 6). For the hybrid CNN-LSTM-Bidirectional model, the authors add the bidirectional LSTM layers
+    learn dependencies in backward direction (Fig 7). For the hybrid CNN-LSTM-Bidirectional model, the authors add the bidirectional LSTM layers
     after the final convolutional layers and before the final dense layers.
@@
 \section{Results and Discussion}
-The DNA sequence dataset used in this paper consists of 66,153 inputs which is divided into training, 
-validation and testing sets with a ratio of 70\%, 10\% and 20\% respectively.
-The training set consists of 46307, and the validation set consists of 6615, and the testing set consists of 13231 samples. 
-The maximum sequence length is 2000, and the vocabulary size is 8972. 
   The proposed models CNN, CNN-LSTM, CNN-Bi-LSTM models are tested by varying different hyperparameters listed in Table I. 
@@
   \subsection{Model Comparison}
-Figure 7 shows the accuracy comparison for all three models for training and testing data using label and k-mer encoding. 
+    Figure 8 shows the accuracy comparison for all three models for training and testing data using label and k-mer encoding. 
     The testing accuracies for label encoding are less when compared to its training accuracy. 
@@
       \includegraphics[width=8cm]{figures/accuracy.png}
-  \caption{Bidirectional LSTM Model Architecture, borrowed from Gunasekaran, et al. (2020)}
+      \caption{Training and Testing Accuracies of all six models, borrowed from Gunasekaran, et al. (2020)}
     \end{figure}
 
-Figure 8 shows the training and validation accuracies for all three models and both encodings. 
-From Figure 8, we can observe that the accuracy of the model remains same after 10 epochs for all models 
+    Figure 9 shows the training and validation accuracies for all three models and both encodings. 
+    From Figure 9, we can observe that the accuracy of the model remains same after 10 epochs for all models 
     except CNN-LSTM using k-mer encoding. This model has unstable accuracy as each epoch increases.
@@
 
-% FIXME: I wanted to use figure* environment to span both columns, but the damn thing is broke
-% It won't stay positioned "here". The documentation discusses figure* but doesn't give much insight
-% into how to fix. So for now, let this span a single column, which makes it look super fooking tiny
-\begin{figure}
-\centering
-  \includegraphics[width=8cm]{figures/train_vrs_valid_accuracies.png}
-  \caption{Training / Validation accuracies for all six models, borrowed from Gunasekaran, et al. (2020)}
-\end{figure}
-
-\subsection{Model Performance Against State of the Art Methods}
-
-The authors compare their accuracy results to other state-of-the-art approaches to 
-virus classification (Fig 9) Their results show that all three of their models show higher accuracy than
-that achieved by Nguyen, et al. (2016), Do, et al. (2020), and Zhang, et al. (2020).
-
-\begin{figure}
-  \centering
-  \includegraphics[width=8cm]{figures/accuracy_compared_to_state_of_art.png}
-  \caption{Gunasekaran, H. et al, compared against state of the art methods, borrowed from Gunasekaran, et al. (2020)}
-\end{figure}
-
-% 3.1 Model comparison :
-% The proposed models are compared with different techniques from [1,2,3] to prove the robustness of the models. 
-% The achieved accuracy is obtained in classifying six classes. 
-% It is observed that all three models have achieved higher accuracies when compared to the state-of-the-art methods.
-
-% Figure 6: Comparison of the models with the state of the art methods.
-
-% Table 4 : Model performance concerning different classes using label and k-mer encoding techniques.
-
-% Table 4 shows the model performance  concerning six different classes using label and k-mer encoding techniques. Based on metrics in Table 4, we can conclude that 
-% ⦁	If class samples are high, CNN with label encoding has high precision.
-% ⦁	If class samples are low, CNN with k-mer encoding has high precision.
-% ⦁	If recall is preferred, k-mer encoding has higher recall for all classes irrespective of the class samples.  
-% ⦁	High sensitivity rate of 99.95% is obtained for class “a” using CNN+Bi-LSTM with label encoding.
-% ⦁	To obtain a high recall and sensitivity for class with more samples, Bi-LSTM with label encoding will be a good choice.
-% ⦁	CNN with label encoding offers high specificity irrespective of the class size.
-% Conclusion :
-% This paper compared three models with label and k-mer encoding techniques.
+    \subsection{Discussion}
+      Figure 10 shows the model performance  concerning six different classes using label and k-mer encoding techniques. 
+      Based on metrics in Figure 10, we make the following observations:
+      \begin{itemize}
+        \item If class samples are high, CNN with label encoding has high precision.
+        \item If class samples are low, CNN with k-mer encoding has high precision.
+        \item If recall is preferred, k-mer encoding has higher recall for all classes irrespective of the class samples.  
+        \item High sensitivity rate of 99.95\% is obtained for class “a” using CNN+Bi-LSTM with label encoding.
+        \item To obtain a high recall and sensitivity for class with more samples, Bi-LSTM with label encoding will be a good choice.
+        \item CNN with label encoding offers high specificity irrespective of the class size.
+      \end{itemize}
       CNN with label encoding outperforms other two models but their testing accuracies are low. 
       K-mer encoding has achieved high validation and testing accuracy. 
-We can conclude that, the best model can be chose not only based on the accuracy of the model but also other metrics such as sensitivity, specificity and precision must be considered.
-% References :
-% ⦁	M.
-F.
-Aslan, M.
-F.
-Unlersen, K.
-Sabanci, and A.
-Durdu, “CNNbased transfer learning-BiLSTM network: a novel approach for COVID-19 infection detection,” Applied Soft Computing, vol.
-98, article 106912, 2021.
+      We can conclude that, the best model can be chose not only based on the accuracy of the model 
+      but also other metrics such as sensitivity, specificity and precision must be considered.
 
-% ⦁	X. Zhang, B. Beinke, B. Al Kindhi, and M. Wiering, “Comparing machine learning algorithms with or without feature extraction for DNA classification,” 2020, http://arxiv.org/abs/2011.00485.
-
-% ⦁	S.
-W.
-I.
-Maalik and S.
-K.
-W.
-Ananta, “Comparation analysis of ensemble technique with boosting (Xgboost) and bagging (Randomforest) for classify splice junction DNA sequence Computational and Mathematical Methods in Medicine 11 category,” Jurnal Penelitian Pos dan Informatika, vol.
-9, no.
-1,pp.
-27–36, 2019.
 \section{Conclusion}
+  The authors' results are promising when compared to existing non machine-learning and machine learning methods for
+  virus classification. The various performance metrics reviewed here show that the right model may depend on the
+  particular use case. If accuracy is the most important metric, then the authors' results show that CNN outperforms
+  CNN-LSTM and CNN-bidrectional LSTM. The high accuracy acheived in this paper demonstrates the effectiveness of 
+  deep learning models in performing complex tasks.
 
@@
   \emph{Computational and mathematical methods in medicine}, 2021.
+\bibitem{IEEEhowto:kopka}
+  S.~Shadab,~M.~T.~Alam~Khan,~N.~A.~Neezi,~S.~Adilina,~and~S.~Shatabda, 
+  “DeepDBP: deep neural networks for identification of DNA-binding proteins,” 
+  \emph{Informatics in Medicine Unlocked}, vol. 19, article 100318, 2020.
+\bibitem{IEEEhowto:kopka}
+  M~.Pop and S~Salzberg. "Bioinformatics challenges of new sequencing technology." 
+  \emph{Trends Genet}. vol. 24, no. 3, 142-149. 2008, doi: 10.1016/j.tig.2007.12.006.
+\bibitem{IEEEhowto:kopka}
+  S.~C.~Pakhrin,~B.~Shrestha,~B.~Adhikari,~and~D.~B.~Kc, 
+  "Deep learning-based advances in protein structure prediction," 
+  \emph{Int. J. Mol. Sci.}, vol. 22, no. 11, 2021, doi: 10.3390/ijms22115553.
+\bibitem{IEEEhowto:kopka}
+  A.~Tasdelen and B.~Sen. "A hybrid CNN-LSTM model for pre-miRNA classification."
+  \emph{Scientific reports}, vol. 11, no. 1, 1-9. 2021.
+\bibitem{IEEEhowto:kopka}
+  J.~Chinju, O.~Matthew,~and~J.~Sahoo. "CNN-LSTM based classification of polo like kinase family of Proteins: An emerging cancer drug target." 
+  \emph{Materials Today: Proceedings}, 2022.
+\bibitem{IEEEhowto:kopka}
+  S.~K.~Sønderby~and~O.~Winther, "Protein Secondary Structure Prediction with Long Short Term Memory Networks," 2014, 
+  [Online]. Available: http://arxiv.org/abs/1412.7828.
 \bibitem{IEEEhowto:kopka}
@@
   2020, http://arxiv.org/abs/2011.00485
+
 \end{thebibliography}
 
+% FIXME: I wanted to use figure* environment to span both columns, but the damn thing is broke
+% It won't stay positioned "here". The documentation discusses figure* but doesn't give much insight
+% into how to fix. So for now, let this span a single column, which makes it look super fooking tiny
+\begin{figure*}[t]
+\centering
+  \includegraphics[width=\textwidth]{figures/train_vrs_valid_accuracies.png}
+  \caption{Training / Validation accuracies for all six models, borrowed from Gunasekaran, et al. (2020)}
+\end{figure*}
+
+\begin{figure*}[t]
+\centering
+  \caption{Performance Metrics for all Models, borrowed from Gunasekaran, et al. (2020)}
+  \includegraphics[width=\textwidth]{figures/performance_metrics.png}
+\end{figure*}
+
 \end{document}

### 2022-04-07 paper_discussion_report/paper_discussion.tex
Fixed some typos in paper discussion report

@@
   The National Centre for Biotechnology Information (NCBI) https://www.ncbi.nlm.nih.gov.
-  They then encode this data using label encoding and kmer encoding. For each encoding type,
+  They then encode this data using label encoding and k-mer encoding. For each encoding type,
   they run three different deep learning models: CNN, CNN-LSTM, and CNN-Bidirectional-LSTM.
@@
   preserving positional information (Fig 4).
-  In the second approach, they used kmer encoding, which generates all kmers from a sequence 
+  In the second approach, they used k-mer encoding, which generates all k-mers from a sequence
   and forms an English-like sentence onto which natural language processing techniques
@@
     \includegraphics[width=8cm]{figures/kmer_encoding.png}
-    \caption{Kmer encoding example, where k=6, borrowed from Gunasekaran, H., et al. (2020)}
+    \caption{K-mer encoding example, where k=6, borrowed from Gunasekaran, H., et al. (2020)}
   \end{figure}
@@
   All three models share the same initial layers: an embedding layer, convolutional layers, and
-  max pooling layers. The raw data is first either label encoded or kmer encoded and then 
+  max pooling layers. The raw data is first either label encoded or k-mer encoded and then
   one-hot encoded to be fed into the embedding layer. The results of the embedding layer
@@
   and specificity by obtaining the confusion matrix for both k-mer and label encoding techniques.
-  The confusion matrix holds the values of true positives (TP), true negatives (TN), false positives (FP), and false negatives(FN). 
+  The confusion matrix holds the values of true positives (TP), true negatives (TN), false positives (FP), and false negatives (FN).
   Based on these values, classification metrics are calculated and compared.

### 2022-05-09 project_discussion_report/main.tex
Added start of project discussion report

\documentclass[journal]{IEEEtran}
\usepackage{tikz}
\usepackage{cite}
\usetikzlibrary{decorations.pathreplacing,angles,quotes}

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
\title{Project Report}

% puts info about authors at bottom of first page
\author{Lawrence~Owusu,~Jordan~Sturtz,~and~Swetha~Chittam~% <-this % stops a space
  \thanks{The authors are graduate students at NCA\&T}% <-this % stops a space
}

% note the % following the last \IEEEmembership and also \thanks -
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!

% The paper headers
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
\markboth{CS851 - Deep Learning: Project Report}%
{Shell \MakeLowercase{\textit{et al.}}: CS851 - Deep Learning: Project Report}

% make the title area
\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
Abstract - clearly state problem, method, and results.
\end{abstract}

% Note that keywords are not normally used for peerreview papers.
% \begin{IEEEkeywords}
% IEEE, IEEEtran, journal, \LaTeX, paper, template.
% \end{IEEEkeywords}

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi

%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Introduction}
  \IEEEPARstart{P}{rotein} structure sequencing remains one of the challenges in the field of computational biology.
    Determination of accurate protein structure is important in developing deep understanding of the
    functions of proteins and their applications of drug and inhibitor discovery and design.

    Protein sequencing begins in the lab using techniques such as mass spectrometry or Edman degradation
    to identify partial sequences of proteins \cite{mann2016rise, edman1949method}. Various techniques exist for using these
    partial sequences to construct the full protein sequence
    \cite{standing2003peptide,perkins1999probability, eng1994approach, yang2021full}.

    In our paper, we develop an approach using deep learning techniques to predict missing gaps in peptide sequences.
    Our approach begins with a partial sequence with gaps of known size. To retrieve the relevant training data,
    we query the National Center For Biotechnology Information (NCBI) Protein Blast server to retrieve the highest matching
    homologous sequences to our partial sequence \cite{ncbi}. We train two models: one to predict peptides in the forward direction and one to predict in the reverse direction.
    We then use both predictions to predict the missing peptides. Our deep learning models use LSTM layers to learn the
    forward and reverse dependencies.

    \subsection{Related Work}
    % Related work should incldue current progress in sequencing, Genome sequencing and single cell sequencing.
    % Should also refer to paper we are using as our basis for our project.
    % Use google scholar to find those who cited our paper and those citations the paper has.

    The history of using Mass Spectrometry to sequence proteins dates to several years ago.
    In this approach, multiple proteases are employed to cleave the same protein sample separately.
    Because the cleavage of protein by the protease generates overlapping peptides, merging the spectral
    pairs of the overlapping peptides consecutively assembles into long contigs from which de novo
    sequences are obtained \cite{liu2014novo}.

    Some researchers on protein sequencing problems focus on identifying proteins or generating full protein sequences
    by comparing partial sequences to known protein databases to infer the missing gaps in mass spectral data
    \cite{eng1994approach, perkins1999probability}. Others focus on developing techniques for de novo protein sequencing,
    which refers to the process of sequencing proteins without the use of a protein database or with minimal use of
    genomic data \cite{standing2003peptide, bandeira2008automated}.

    Over the last 10 years, de novo protein sequencing has been researched
    extensively in computational proteomics and have been used successfully to deduce peptide sequence
    of un-sequenced organisms, antibodies and post-translastionally modified peptides
    \cite{ma2012novo,veltman2012novo, bandeira2007spectral}. For example,
    Mai et al. reported that their assembling algorithm, Multiple Contigs and Scaffolding
    successfully assembles the de novo identified peptides in contig-scaffold fashion, resulting in
    100\% coverage and 98.69-100\% accuracy on three proteins and replications. The Multiple Contigs
    and Scaffolding algorithm has provided robust and accurate software for full-length protein sequencing
    after de novo identification of peptides \cite{mai2022highly}

    Similarly, Yang et al. reported 100\% accuracy for full-length de novo sequencing
    for light chains of Herceptin and bovine serum albumin (BSA) when their proposed method was applied to de novo
    sequencing of bovine serum albumin (BSA) and monoclonal antibody Herceptin. However, the accuracy marginally dropped to
    99.7\% for the heavy chains of Herceptin \cite{yang2021full}.

    Protein sequencing is only one type of sequencing problem in biology. Related types of sequencing problems include
    single-cell protein sequencing or whole genomic sequencing \cite{wang2015advances,pareek2011sequencing}.

\section{Method}

%   Method: Should have two subsctions (1) Data discussion (preprocessing, where we got it from etc), (2) LSTM / CNN model
  \subsection{Data Collection}
    For our initial project, we used as a target sequence the light chain of alemtuzumab referenced in Liu, et al.
    \cite{liu2014novo}. We manually removed the gaps produced by Liu et al.'s tandem mass spectrometry approach, then
    entered that sequence into NCBI's Protein Blast Server to retrieve the closest matching homologous sequences \cite{ncbi}.
    We used the top ten homologous sequences as training data. The closest matching homologous sequences in our training
    data range from 89.32\% to 82.52\% similarity.

  \subsection{Data Preprocessing}
  We generated from our homologous sequences all kmers. A Kmer is a any substring of length K. For our models, we chose a
  kmer-length of 5. Each kmer represents a single input. The output of each associated kmer is the next character in the sequence.
  Thus, for instance, if one of our homologous sequences contains the substring "DIQM", then a single training instance would be
  the input-output tuple (DIQ, M).

  From our ten homologous sequences, we generated 2106 number of input-output pairs with an input length of 5.
  The kmer length will represent the timesteps in our LSTM layer.
  To be fed into a neural network, the data must be encoded numerically, so we assigned integer labels to each character.
  The output of each training instance is a
  one-hot encoded vector to represent the target classes. Since we are training two models, one for forward prediction
  and one for reverse prediction, we generate the same training data in both the reverse and forward directions.

  The samples of forward and reverse input-output pairs are divided into train and validation sets with the ratio of
  90\% and 10\% respectively. The validation set is used to perform hyperparameter tuning.

\subsection{Data Normalization}
    The input pairs are normalized by dividing with number of classes to obtain a range between [0, 1].
    The number of classes are determined by extracting all the unique characters from the input sequences.
    This data normalization technique is applied to forward, reverse and test data input-output pairs.
    Normalization helps speed the training process.

  \subsection{Model Overview}

    For our sequence prediction task, we built four deep learning models combining convolutional neural
    networks (CNN) with long short term memory (LSTM). The CNN layer helps with automatic feature extraction, in particular
    if there are patterns within each kmer that can be extracted for better predictions. The LSTM layer learns
    the dependencies in the sequence data to predict the correct output. For our four hybrid models, we also
    added two dense hidden layers with ReLu activation and a final output layer with softmax activation.

    Our four hybrid models are LSTM, CNN-LSTM, Bi-LSTM, and CNN-Bi-LSTM.
    The purpose of using four models is to compare their results.

    The workflow for the proposed work is shown in Figure 1.

    % We added two extra dense hidden layers with ReLu activation after our LSTM layer and a final output layer, using softmax activation.

    % Our four hybrid models
    % We built four hybrid models for the de novo sequence predictions:
    % LSTM, CNN-LSTM, Bi-LSTM and CNN-Bi-LSTM.
    % For all four models, we used the same preprocessing techniques.
    % The CNN layer is used as the feature extraction stage, and it is given as the input to
    % the LSTM/Bi-LSTM layers which passes its outputs to two hidden dense relu-activated layers and finally
    % the output layer. The output layer uses softmax activation. The loss function for all the models
    % is cross entropy.

    \begin{figure}[h]
      \centering
      \includegraphics[width=8cm]{Model_Diagram.JPG}
      \caption{Model diagram for multi-classification of sequence data}
    \end{figure}

    \subsubsection{CNN}
    CNN is a powerful deep learning technique for feature extraction.
    CNN is often used on 2-dimensional and 3-dimensional datasets, e.g. image or video datasets.
    We use a 1-dimensional convolutional layer that convolves the kmers to extract any meaningful
    features. The convolutional layer has 128 filters with a kernel size of 3, so minimally the kmer-length cannot be
    smaller than 3. The feature maps from this layer forms as input to the LSTM or Bi-LSTM layer.

    \subsubsection{LSTM}
    Characteristic of the LSTM architecture is a set of chained together cells called "memory blocks". The number
    of memory blocks in the chain equals the length of the timesteps in our target dataset (Fig 2).
    Each memory cell has three gating units (forget, input and output gates) which
    conditionally regulate how information flows into and out of the memory block (Fig 3). Intuitively,
    the "forget" gate can be viewed as a step where irrelevant information from the hidden state
    is first "forgotten" before being passed to the input gate, which constructs the new cell state.
    Before that cell state can output its value to the next memory cell, it is filtered again through
    an output gate that decides what values to keep internal to the memory cell (i.e. the hidden state)
    and what to output to the next memory cell or final output layer.
    The LSTM layer is helpful, therefore, for learning the important sequence dependencies for performing
    sequence predictions.

    \subsubsection{Bidirectional LSTM}
        Bidirectional LSTM is a modification of the LSTM architecture that permits the model to learn both the
        forward and reverse dependencies. The way it acheives this is to have two LSTM layers, one consuming
        the timesteps in the forward direction and the other consuming the timesteps in the reverse direction.
        The two layers then merge their results to produce the final output.

    \begin{figure}[h]
      \centering
      \includegraphics[width=8cm]{bidirectional-lstm.jpeg}
      \caption{Bidirectional LSTM Architecture}
    \end{figure}

    \begin{figure}[h]
    \centering
    \tikzstyle{place}=[rectangle,draw, thick, minimum width=30, minimum height=10]
    \begin{tikzpicture}
        \node at (0,0) [place] (first) {Memory Cell};
        \node at (3,0) [place] (second) {Memory Cell};
        \node at (6,0) [place] (third) {Memory Cell};
        % \draw [->] (0,0) -- (first.west);
        \draw [->] (first.east) -- (second.west);
        \draw [->] (second.east) -- (third.west);
        % \draw [->] (third.east) -- (10,0);
        \draw[decoration={brace,mirror,raise=10pt},decorate](first.south west) -- node[below=16pt] {Number of cells = number of time steps} (third.south east);
    \end{tikzpicture}
    \caption{Illustration of recurrently chained memory cells}
    \end{figure}

    \begin{figure}[h]
      \centering
      \includegraphics[width=8cm]{jordan_lstm_architecture.png}
      \caption{LSTM architecture inside a memory cell}
    \end{figure}

    \section{Model Architectures}

%   Model architecture: CNN / LSTM. Create diagram showing overall architecture. Discuss hyperparameters of model.

  \subsection{CNN}
  \subsection{LSTM}
  \subsection{LSTM-CNN}
  \subsection{BiLSTM-CNN}

\section{Experimental Results}
  Experimental results: Implementation process, tuning hyperparameters,
  show predictions in amino acid gaps
  (deploy the model and show its accuracy on a particular instance)
  \subsection{Tuning hyperparameters}
  \subsection{Accuracy/loss/confusion matrix/whatever}
  \subsection{Predictions on De Novo Sequence}

\section{Conclusion}
  Conclusion: Summarize project

\bibliographystyle{IEEEtran}
\bibliography{main}

% \begin{thebibliography}{10}

% \bibitem{cnn-hybrid}
%   H.~Gunasekaran, K.~Ramalakshmi, A.~Rex~Macedo~Arokiaraj, S.~Deepa~Kanmani, C.~Venkatesan, and C.~Suresh~Gnana~Dhas.
%   "Analysis of DNA Sequence Classification Using CNN and Hybrid Models."
%   \emph{Computational and mathematical methods in medicine}, 2021.
% \bibitem{IEEEhowto:kopka}
%   S.~Shadab,~M.~T.~Alam~Khan,~N.~A.~Neezi,~S.~Adilina,~and~S.~Shatabda,
%   “DeepDBP: deep neural networks for identification of DNA-binding proteins,”
%   \emph{Informatics in Medicine Unlocked}, vol. 19, article 100318, 2020.
% \bibitem{IEEEhowto:kopka}
%   M~.Pop and S~Salzberg. "Bioinformatics challenges of new sequencing technology."
%   \emph{Trends Genet}. vol. 24, no. 3, 142-149. 2008, doi: 10.1016/j.tig.2007.12.006.
% \bibitem{IEEEhowto:kopka}
%   S.~C.~Pakhrin,~B.~Shrestha,~B.~Adhikari,~and~D.~B.~Kc,
%   "Deep learning-based advances in protein structure prediction,"
%   \emph{Int. J. Mol. Sci.}, vol. 22, no. 11, 2021, doi: 10.3390/ijms22115553.
% \bibitem{IEEEhowto:kopka}
%   A.~Tasdelen and B.~Sen. "A hybrid CNN-LSTM model for pre-miRNA classification."
%   \emph{Scientific reports}, vol. 11, no. 1, 1-9. 2021.
% \bibitem{IEEEhowto:kopka}
%   J.~Chinju, O.~Matthew,~and~J.~Sahoo. "CNN-LSTM based classification of polo like kinase family of Proteins: An emerging cancer drug target."
%   \emph{Materials Today: Proceedings}, 2022.
% \bibitem{IEEEhowto:kopka}
%   S.~K.~Sønderby~and~O.~Winther, "Protein Secondary Structure Prediction with Long Short Term Memory Networks," 2014,
%   [Online]. Available: http://arxiv.org/abs/1412.7828.
% \bibitem{IEEEhowto:kopka}
%   N.~G.~Nguyen,~V.~A.~Tran,~D.~L.~Ngo et al., “DNA sequence classification by convolutional neural network,” \emph{Journal of Bio-
%   medical Science and Engineering}, vol. 9, no. 5, pp. 280–286, 2016.
% \bibitem{IEEEhowto:kopka}
%   D.~T.~Do~and~N.~Q.~K.~Le, “Using extreme gradient boosting to identify origin of replication in Saccharomyces cerevisiae via
%   hybrid features,” \emph{Genomics}, vol. 112, no. 3, pp. 2445–2451,2020
% \bibitem{IEEEhowto:kopka}
%   X.~Zhang,~B.~Beinke,~B.~Al~Kindhi,~and~M.~Wiering, “Comparing machine learning algorithms with or without feature extraction for DNA classification"
%   2020, http://arxiv.org/abs/2011.00485

% \end{thebibliography}
\end{document}

### 2022-05-09 lstm.py

@@ -70,33 +70,8 @@ def save_result(test_accuracy, history, model, secs_to_train):
     ax.set_title(f'Accuracy')  # Add a title to the axes.
     ax.xaxis.set_major_locator(MaxNLocator(integer=True))
 
-    # FIXME: Was trying to add the model parameters to the figure but fuck it
-    # notes = {
-    #     "test_accuracy": test_accuracy,
-    #     "epochs_actually_trained": epoch_length,
-    # }
-
-    # notes.update(parameters)
-    # note = "\n".join(f"{name:<{pad_name}}: {value:>{pad_value}}" for name, value in notes.items())
-
-    # ax.annotate(note,
-    #             xy = (1.1, 0.5),
-    #             xycoords='axes fraction',
-    #             ha='left',
-    #             va="center",
-    #             fontsize=10)
-
     ax.legend();  # Add a legend.
     filename = f"{now}"
-
-    # FIXME: This shit sucks. The actual method to call depends on the backend used for the UI. This only works if backend
-    # is QT. Really stupid that it's written to be backend specific. Should be a universal call regardless of backend to
-    # ask the screen to render it (and save it) maximized.
-    # Since the size of the saved image is based on "maximizing" which is dependent on the particular machine that runs it,
-    # this code produces different size images for different machines, which also sucks. I'm too lazy to figure out how to
-    # Resize this thing appropriately
-    # manager = plt.get_current_fig_manager()
-    # manager.window.showMaximized()
     fig.savefig(os.path.join(RESULTS_PATH, f"{filename}_accuracy.png"))
 
     parameters = {
@@ -112,7 +87,6 @@ def save_result(test_accuracy, history, model, secs_to_train):
         "optimizer": OPTIMIZER,
         "output_activation_func": OUTPUT_ACTIVATION_FUNC,
         "hidden_layer_activation_func": HIDDEN_LAYER_ACTIVATION_FUNC,
-        # "training/validation/test split": f"{int(TRAINING_SPLIT * 100)}/{int(VALIDATION_SPLIT* 100)}/{int(TEST_SPLIT * 100)}",
         "patience_threshold": PATIENCE_THRESHOLD,
         "num_classes": NUM_CLASSES,
         "batch_size": BATCH_SIZE,
@@ -121,10 +95,8 @@ def save_result(test_accuracy, history, model, secs_to_train):
     result = {
         "test_accuracy": test_accuracy,
[…]

### 2022-05-09 lstm.py
Updated lstm.py to use different models for builds, and to generate loss curves

@@ -4,8 +4,8 @@ import json
 import numpy as np
 import os
 from datetime import datetime
+from enum import Enum
 
-from keras.models import Sequential
 from tensorflow import keras
 from tensorflow.keras import layers
 from keras.utils import np_utils
@@ -27,7 +27,7 @@ np.random.seed(7)
 # =============================================================================
 
 # Path to save results
-RESULTS_PATH = os.path.join(os.getcwd(), "new_approach_results")
+RESULTS_PATH = os.path.join(os.getcwd(), "new_new_approach_results")
 
 # Model parameters
 KMER_LENGTH = 5
@@ -56,33 +56,44 @@ INT_TO_CHAR = {}
 
 # =============================================================================
 # Save what we want from the results
-def save_result(test_accuracy, history, model, secs_to_train):
+def save_result(title, test_accuracy, history, model, model_type="Default"):
     now = datetime.now().strftime("%H-%M-%S")
     epoch_length = len(history["epochs_to_patience"])
     epoch_axis = list(range(1, epoch_length + 1))
 
-    # Note that even in the OO-style, we use `.pyplot.figure` to create the Figure.
     fig, ax = plt.subplots()
-    ax.plot(epoch_axis, history["accuracy"], label='Training')
-    ax.plot(epoch_axis, history["val_accuracy"], label='Validation')
+    ax.plot(epoch_axis, history["loss"], label='Training')
+    ax.plot(epoch_axis, history["val_loss"], label='Validation')
     ax.set_xlabel('Epochs')  # Add an x-label to the axes.
-    ax.set_ylabel('Accuracy')  # Add a y-label to the axes.
-    ax.set_title(f'Accuracy')  # Add a title to the axes.
+    ax.set_ylabel('Loss')  # Add a y-label to the axes.
+    ax.set_title(f'Loss on {model_type}')  # Add a title to the axes.
     ax.xaxis.set_major_locator(MaxNLocator(integer=True))
 
     ax.legend();  # Add a legend.
-    filename = f"{now}"
+    filename = f"{now}{title}_{model_type}"
+    fig.savefig(os.path.join(RESULTS_PATH, f"{filename}_loss.png"))
+
+    fig, ax = plt.subplots()
[…]

### 2022-05-09 project_discussion_report/main.tex
Added final writeup for Project Report

@@
 \begin{abstract}
-Abstract - clearly state problem, method, and results.
+  De novo protein sequencing using mass spectrometry or Edman degradation
+  suffers from the problem of producing only partial sequences of target proteins.
+  We explore the possibility of using deep learning techniques to perform the 
+  task of predicting gaps in partially sequenced proteins. Our method involves
+  querying the NCBI Protein Blast server for closest matching homologous sequences
+  to a partial sequence, then training two deep learning models to predict
+  in the forward and reverse directions the missing gaps. Our results show
+  high accuracy (98-99\%) in filling gaps of a particular partially sequenced 
+  de novo sequence of the light chain of alemtuzumab.
+
 \end{abstract}
@@
     we query the National Center For Biotechnology Information (NCBI) Protein Blast server to retrieve the highest matching
-    homologous sequences to our partial sequence \cite{ncbi}. We train two models: one to predict peptides in the forward direction and one to predict in the reverse direction.
-    We then use both predictions to predict the missing peptides. Our deep learning models use LSTM layers to learn the
-    forward and reverse dependencies.
+    homologous sequences to our partial sequence \cite{ncbi}.We then train and deploy two models: one to learn dependencies
+    in the forward direction and another to learn dependencies in the reverse direction. We combine the two 
+    models' predictions to fill the gaps in the de novo protein sequence.
+
+    We train and evaluate four competing hybrid deep learning models combining LSTM and CNN layers.     
+    For the purpose of this project, we tested our model against only a single 
+    target sequence: the light chain of alemtuzumab. Our results show high accuracies (98-99\%) for all four of our model architectures,
+    with our Bi-LSTM-CNN model outperforming the others at 99.04\%.
 
     \subsection{Related Work}
-    % Related work should incldue current progress in sequencing, Genome sequencing and single cell sequencing.
-    % Should also refer to paper we are using as our basis for our project.
-    % Use google scholar to find those who cited our paper and those citations the paper has.
 
-    The history of using Mass Spectrometry to sequence proteins dates to several years ago.
+    Mass spectrometry continues to be a dominant method to perform protein sequencing \cite{mann2016rise}.
     In this approach, multiple proteases are employed to cleave the same protein sample separately.
@@
     extensively in computational proteomics and have been used successfully to deduce peptide sequence
-    of un-sequenced organisms, antibodies and post-translastionally modified peptides
+    of un-sequenced organisms, antibodies and post-translationally modified peptides
     \cite{ma2012novo,veltman2012novo, bandeira2007spectral}. For example,
@@
 
-%   Method: Should have two subsctions (1) Data discussion (preprocessing, where we got it from etc), (2) LSTM / CNN model
   \subsection{Data Collection}
@@
   \subsection{Data Preprocessing}
-  We generated from our homologous sequences all kmers. A Kmer is a any substring of length K. For our models, we chose a
-  kmer-length of 5. Each kmer represents a single input. The output of each associated kmer is the next character in the sequence.
-  Thus, for instance, if one of our homologous sequences contains the substring "DIQM", then a single training instance would be
-  the input-output tuple (DIQ, M).
+    We generated from our homologous sequences all kmers. A kmer is a any substring of length K. For our models, we chose a
+    kmer-length of 5. Each kmer represents a single input, and the output of each associated kmer is the next character in the sequence.
+    Thus, for instance, if one of our homologous sequences contains the substring "DIQMSQ", then a single training instance would be
+    the input-output tuple (DIQMS, Q).
 
     From our ten homologous sequences, we generated 2106 number of input-output pairs with an input length of 5.
-  The kmer length will represent the timesteps in our LSTM layer.
+    The kmer length represents the timesteps in our LSTM layer.
     To be fed into a neural network, the data must be encoded numerically, so we assigned integer labels to each character.
@@
 
-  The samples of forward and reverse input-output pairs are divided into train and validation sets with the ratio of
-  90\% and 10\% respectively. The validation set is used to perform hyperparameter tuning.
+    The samples of forward and reverse input-output pairs are divided into training and validation sets with the ratio of
+    90\% and 10\% respectively. The validation set is used to prevent overfitting.
 
@@
 
-  \subsection{Model Overview}
+\subsection{Model Selection}
 
     For our sequence prediction task, we built four deep learning models combining convolutional neural
-    networks (CNN) with long short term memory (LSTM). The CNN layer helps with automatic feature extraction, in particular
-    if there are patterns within each kmer that can be extracted for better predictions. The LSTM layer learns
-    the dependencies in the sequence data to predict the correct output. For our four hybrid models, we also
-    added two dense hidden layers with ReLu activation and a final output layer with softmax activation.
+    networks (CNN) with long short term memory (LSTM). The CNN layer in theory helps with automatic feature extraction, in particular
+    if there are patterns within each kmer that can be extracted for better predictions. The LSTM layer is the 
+    core of our models, since LSTM is useful for learning sequence dependencies in our input data.
+    For our four hybrid models, we also added two dense hidden layers with ReLu activation and a
+    final output layer with softmax activation.
 
-    Our four hybrid models are LSTM, CNN-LSTM, Bi-LSTM, and CNN-Bi-LSTM.
-    The purpose of using four models is to compare their results.
-
-    The workflow for the proposed work is shown in Figure 1.
-
-    % We added two extra dense hidden layers with ReLu activation after our LSTM layer and a final output layer, using softmax activation.
-
-    % Our four hybrid models
-    % We built four hybrid models for the de novo sequence predictions:
-    % LSTM, CNN-LSTM, Bi-LSTM and CNN-Bi-LSTM.
-    % For all four models, we used the same preprocessing techniques.
-    % The CNN layer is used as the feature extraction stage, and it is given as the input to
-    % the LSTM/Bi-LSTM layers which passes its outputs to two hidden dense relu-activated layers and finally
-    % the output layer. The output layer uses softmax activation. The loss function for all the models
-    % is cross entropy.
+    It is not clear a priori whether LSTM or Bi-LSTM would be more effective in our prediction task; nor is it clear
+    whether the addition of the convolutional layer would improve our results. We thus opted to evaluate and compare
+    four hybrid models: LSTM, CNN-LSTM, Bi-LSTM, and CNN-Bi-LSTM.
+    The workflow for the proposed approach is shown in Figure~\ref{fig:model_diagram}.
 
@@
       \centering
-      \includegraphics[width=8cm]{Model_Diagram.JPG}
-      \caption{Model diagram for multi-classification of sequence data}
+      \includegraphics[width=8cm]{figures/Model_Diagram.JPG}
+      \caption{Model Diagram for Multi-Classification of Sequence Data}
+      \label{fig:model_diagram}
     \end{figure}
@@
     \subsubsection{LSTM}
-    Characteristic of the LSTM architecture is a set of chained together cells called "memory blocks". The number
-    of memory blocks in the chain equals the length of the timesteps in our target dataset (Fig 2).
-    Each memory cell has three gating units (forget, input and output gates) which
-    conditionally regulate how information flows into and out of the memory block (Fig 3). Intuitively,
-    the "forget" gate can be viewed as a step where irrelevant information from the hidden state
+      The LSTM architecture consists of a set of chained together cells called "memory blocks". The number
+      of memory blocks in the chain equals the length of the timesteps in our target dataset.
+      Each memory cell has three gating units (forget, input, and output gates) which
+      conditionally regulate how information flows into and out of the memory block (Figure~\ref{fig:lstm_architecture}). 
+      Intuitively, the "forget" gate can be viewed as a step where irrelevant information from the hidden state
       is first "forgotten" before being passed to the input gate, which constructs the new cell state.
@@
 
-    \subsubsection{Bidirectional LSTM}
-        Bidirectional LSTM is a modification of the LSTM architecture that permits the model to learn both the
-        forward and reverse dependencies. The way it acheives this is to have two LSTM layers, one consuming
+      \begin{figure}[h]
+        \centering
+        \includegraphics[width=8cm]{figures/jordan_lstm_architecture.png}
+        \caption{LSTM Architecture Inside a Memory Cell}
+        \label{fig:lstm_architecture}
+      \end{figure}
+
+    \subsubsection{Bi-LSTM}
+      Bidirectional LSTM (Bi-LSTM) is a modification of the LSTM architecture that permits the model to learn both the
+      forward and reverse dependencies. Bi-LSTM uses two LSTM layers, one consuming
       the timesteps in the forward direction and the other consuming the timesteps in the reverse direction.
-        The two layers then merge their results to produce the final output.
+      The two layers then merge their results to produce the final output. See Figure~\ref{fig:bidirectional_diagram}
+      for an illustration.
 
@@
         \centering
-      \includegraphics[width=8cm]{bidirectional-lstm.jpeg}
+        \includegraphics[width=8cm]{figures/bidirectional-lstm.jpeg}
         \caption{Bidirectional LSTM Architecture}
+        \label{fig:bidirectional_diagram}
+      \end{figure}
+
+\section{Overall Model Architecture}
+
+  All four of our hybrid models share common features. If there is a convolutional
+  layer, it is the first layer after the input layer. After that,
+  the next layer is either the LSTM or Bi-LSTM layer. After that,
+  the results are passed to two hidden densely connected layer with ReLu activation,
+  which pass their results to the final dense layer with
+  softmax activation. The number of neurons in the final output layer
+  equals the number of our target classes. See Figure~\ref{fig:model_architectures}.
+
+  \begin{figure}[h]
+    \centering
+    \includegraphics[width=8cm]{figures/model_architectures.png}
+    \caption{Deep Learning Model Architecture}
+    \label{fig:model_architectures}
+  \end{figure}
+
+\section{Experimental Results}
+  \subsection{Experimental Model Architectures}
+
+    We initially experimented with a simple input-LSTM-output architecture. By itself, a single LSTM layer acheived high
+    accuracy with minimal amount of training. We attribute this largely to the quality of our
+    training dataset. Indeed, all of the training instances had between 89.32\% to 82.52\% similarity
+    match to our target sequence. The lower the similarity between our target sequence and the training
+    data, the lower our expected accuracy and thus the greater the difficulty in the sequencing task.
+
+    We added two hidden Dense layers after the initial LSTM layer and the accuracy improved marginally.
+    In subsequent model trials, we kept the two hidden layers since they showed marginal improvement
+    over leaving them out.
+
+    We settled on four competing architectures to run: LSTM, CNN-LSTM, BiLSTM, and CNN-BiLSTM.
+    Their testing accuracies, final loss values, and final validation loss values are displayed
+    in Table~\ref{tab:model_results}.
+
+    Our models all exhibited slight overfitting, though not enough to be significant.
+    Figures~\ref{fig:lstm_accuracy},~\ref{fig:cnn_accuracy},~\ref{fig:bilstm_accuracy},~and~\ref{fig:cnn_bilstm_accuracy}
+    plot training accuracy against validation accuracy for our four models, and 
+    Figures~\ref{fig:lstm_loss},~\ref{fig:cnn_loss},~\ref{fig:bilstm_loss},~and~\ref{fig:cnn_bilstm_loss}
+    plot training loss versus validation loss for all four models.
+
+    \begin{table}[h!]
+      \begin{center}
+        \begin{tabular}{l|l|l|l} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
+          & \textbf{Test Accuracy} & \textbf{Training Loss} & \textbf{Validation Loss}\\
+          \hline
+          LSTM        & $98.56\%$ & $0.0964$ & $0.4546$ \\
+          Bi-LSTM     & $98.09\%$ & $0.1282$ & $0.6051$ \\
+          CNN-LSTM    & $98.56\%$ & $0.0948$ & $0.5604$ \\
+          Bi-CNN-LSTM & $99.04\%$ & $0.0884$ & $0.6719$ \\
+        \end{tabular}
+        \caption{Comparing Different Model Architectures}
+        \label{tab:model_results}
+      \end{center}
+    \end{table}
+
+    \begin{figure}[h]
+      \centering
+      \includegraphics[width=8cm]{figures/01-58-36Forward_LSTM With Dense Layers_accuracy.png}
+      \caption{LSTM Accuracy}
+      \label{fig:lstm_accuracy}
     \end{figure}
@@
       \centering
-    \tikzstyle{place}=[rectangle,draw, thick, minimum width=30, minimum height=10]
-    \begin{tikzpicture}
-        \node at (0,0) [place] (first) {Memory Cell};
-        \node at (3,0) [place] (second) {Memory Cell};
-        \node at (6,0) [place] (third) {Memory Cell};
-        % \draw [->] (0,0) -- (first.west);
-        \draw [->] (first.east) -- (second.west);
-        \draw [->] (second.east) -- (third.west);
-        % \draw [->] (third.east) -- (10,0);
-        \draw[decoration={brace,mirror,raise=10pt},decorate](first.south west) -- node[below=16pt] {Number of cells = number of time steps} (third.south east);
-    \end{tikzpicture}
-    \caption{Illustration of recurrently chained memory cells}
+      \includegraphics[width=8cm]{figures/02-07-24Forward_CNN LSTM With Dense Layers_accuracy.png}
+      \caption{CNN LSTM Accuracy}
+      \label{fig:cnn_accuracy}
     \end{figure}
@@
       \centering
-      \includegraphics[width=8cm]{jordan_lstm_architecture.png}
-      \caption{LSTM architecture inside a memory cell}
+      \includegraphics[width=8cm]{figures/02-31-57Forward_Bi-LSTM With Dense Layers_accuracy.png}
+      \caption{Bi-LSTM Accuracy}
+      \label{fig:bilstm_accuracy}
     \end{figure}
 
-    \section{Model Architectures}
+    \begin{figure}[h]
+      \centering
+      \includegraphics[width=8cm]{figures/02-34-46Forward_CNN Bi-LSTM With Dense Layers_accuracy.png}
+      \caption{CNN Bi-LSTM Accuracy}
+      \label{fig:cnn_bilstm_accuracy}
+    \end{figure}
 
-%   Model architecture: CNN / LSTM. Create diagram showing overall architecture. Discuss hyperparameters of model.
+    \begin{figure}[h]
+      \centering
+      \includegraphics[width=8cm]{figures/01-58-36ForwardLSTM With Dense Layers_loss.png}
+      \caption{LSTM Loss}
+      \label{fig:lstm_loss}
+    \end{figure}
 
-  \subsection{CNN}
-  \subsection{LSTM}
-  \subsection{LSTM-CNN}
-  \subsection{BiLSTM-CNN}
+    \begin{figure}[h]
+      \centering
+      \includegraphics[width=8cm]{figures/02-07-24ForwardCNN LSTM With Dense Layers_loss.png}
+      \caption{CNN LSTM Loss}
+      \label{fig:cnn_loss}
+    \end{figure}
 
-\section{Experimental Results}
-  Experimental results: Implementation process, tuning hyperparameters,
-  show predictions in amino acid gaps
-  (deploy the model and show its accuracy on a particular instance)
-  \subsection{Tuning hyperparameters}
-  \subsection{Accuracy/loss/confusion matrix/whatever}
+    \begin{figure}[h]
+      \centering
+      \includegraphics[width=8cm]{figures/02-31-57Forward_Bi-LSTM With Dense Layers_loss.png}
+      \caption{Bi-LSTM Loss}
+      \label{fig:bilstm_loss}
+    \end{figure}
+
+    \begin{figure}[h]
+      \centering
+      \includegraphics[width=8cm]{figures/02-34-46Forward_CNN Bi-LSTM With Dense Layers_loss.png}
+      \caption{CNN Bi-LSTM Loss}
+      \label{fig:cnn_bilstm_loss}
+    \end{figure}
+
+  \subsection{Hyperparameters}
+
+    Our hyperparameters were chosen through manual trial-and-error. We did not implement
+    any regularization techniques, nor did we use our validation set or cross validation to tune 
+    any hyperparameters. All of our results for our different model architectures were generating
+    high accuracy (>98\%) on our single de novo sequence. In the future, we would like to experiment
+    more with tuning hyperparameters in particular if the task becomes harder for different starting
+    de novo sequences and therefore different training datasets. Table~\ref{tab:hyperparameters}
+    displays our chosen hyperparameters.
+
+    \begin{table}[h!]
+      \begin{center}
+        \caption{Hyperparameters}
+        \label{tab:hyperparameters}
+        \begin{tabular}{l|l|l} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
+          & \textbf{Hyperparameter} & \textbf{Value} \\
+          \hline
+          \textbf{Shared} & Kmer Length                      & 5 \\
+                          & Epochs                           & 100 \\
+                          & First Dense Layer Neurons        & 128 \\
+                          & Second Dense Layer Neurons       & 64 \\
+                          & Cost Function                    & Categorical Cross Entropy \\
+                          & Optimizer                        & ADAM \\
+                          & Output Activation Function       & Softmax \\
+                          & Dense Layer Activation Function  & ReLu \\
+                          & Batch Size                       & 64 \\
+          \hline
+          \textbf{CNN}    & Number of Convolutional Filters  & 128 \\
+                          & Convolutional Filter Size        & 3 \\
+          \hline
+          \textbf{LSTM}   & Number of LSTM Neurons           & 256 \\
+        \end{tabular}
+      \end{center}
+    \end{table}
+
   \subsection{Predictions on De Novo Sequence}
+    To perform the task of deploying our models to predict gaps in our de novo sequence, 
+    we trained two models: one to learn the dependencies in the forward direction and
+    one to learn the dependencies in the reverse direction. This was necessary to deploy 
+    our models to predict gaps in the beginning or end of the target sequence. 
+    For example, if our target sequence begins with "DIQMSP" and we want to predict "DIQ", then
+    we cannot predict from the forward direction, since "DIQ" is the absolute start of 
+    the sequence. We must therefore be able to predict "DIQ" given the sequence
+    ahead of it. Similarly, we must predict in the forward direction for gaps at the absolute end of the sequence.
+
+    This use of two models implies a potential disagreement: one model may predict
+    a different value than the other for those middle gaps with enough
+    preceding values to create a valid input. See Figure~\ref{fig:forward_reverse_predictions}
+    for an example of two different predictions from one sample run.
+
+    To resolve any disagreement between the two models, 
+    we opted to choose the prediction with the highest corresponding associated
+    probability. By associated probability, we mean the output value
+    of the output class with the highest probability.
+    Since softmax outputs a probability distribution over the target classes,
+    the prediction with the higher associated probability ought to be 
+    preferred since its higher probability represents the "certainty"
+    the model has in its prediction.
+
+    With this simple algorithm, we deployed our dual models to predict
+    gaps in our de novo sequence, achieving 100\% accuracy for the test gaps.
+    See Figure~\ref{fig:full_predictions}.
+
+    \begin{figure}[h]
+      \centering
+      \includegraphics[width=8cm]{figures/forward_reverse_predictions.png}
+      \caption{Illustration of Conflicting Predictions}
+      \label{fig:forward_reverse_predictions}
+    \end{figure}
+
+    \begin{figure}[h]
+      \centering
+      \includegraphics[width=8cm]{figures/full_predictions.png}
+      \caption{Illustration of Complete Predictions}
+      \label{fig:full_predictions}
+    \end{figure}
 
 \section{Conclusion}
-  Conclusion: Summarize project
+  Our project aims to explore the possibility of using deep learning techniques to 
+  solve an otherwise hard problem of predicting gaps in partial peptide de novo 
+  sequences. We have demonstrated that with various LSTM-based deep learning models
+  it is possible to achieve high accuracy on peptide sequences with gaps that have high 
+  similarity scores in known protein databases.
+
+  The approach we describe here generalizes to any other de novo sequence and in fact
+  other sequencing problems beyond protein sequencing. Further research is needed to 
+  explore the challenges and limitations of this approach, in particular for
+  de novo sequences for which the highest similarity score in known protein databases 
+  is comparatively low. Moreover, our approach relies on an assumption that gap 
+  lengths in partial sequences are known in advance. In practice, partial sequence
+  gaps may be unknown. It may be possible to solve this problem by employing
+  forward and reverse predictions to generate overlapping sections or to use other 
+  data features (for instance, known total mass of the gap) to successfully predict
+  gaps of unknown length.
 
@@
 
-% \begin{thebibliography}{10}
-
-% \bibitem{cnn-hybrid}
-%   H.~Gunasekaran, K.~Ramalakshmi, A.~Rex~Macedo~Arokiaraj, S.~Deepa~Kanmani, C.~Venkatesan, and C.~Suresh~Gnana~Dhas.
-%   "Analysis of DNA Sequence Classification Using CNN and Hybrid Models."
-%   \emph{Computational and mathematical methods in medicine}, 2021.
-% \bibitem{IEEEhowto:kopka}
-%   S.~Shadab,~M.~T.~Alam~Khan,~N.~A.~Neezi,~S.~Adilina,~and~S.~Shatabda,
-%   “DeepDBP: deep neural networks for identification of DNA-binding proteins,”
-%   \emph{Informatics in Medicine Unlocked}, vol. 19, article 100318, 2020.
-% \bibitem{IEEEhowto:kopka}
-%   M~.Pop and S~Salzberg. "Bioinformatics challenges of new sequencing technology."
-%   \emph{Trends Genet}. vol. 24, no. 3, 142-149. 2008, doi: 10.1016/j.tig.2007.12.006.
-% \bibitem{IEEEhowto:kopka}
-%   S.~C.~Pakhrin,~B.~Shrestha,~B.~Adhikari,~and~D.~B.~Kc,
-%   "Deep learning-based advances in protein structure prediction,"
-%   \emph{Int. J. Mol. Sci.}, vol. 22, no. 11, 2021, doi: 10.3390/ijms22115553.
-% \bibitem{IEEEhowto:kopka}
-%   A.~Tasdelen and B.~Sen. "A hybrid CNN-LSTM model for pre-miRNA classification."
-%   \emph{Scientific reports}, vol. 11, no. 1, 1-9. 2021.
-% \bibitem{IEEEhowto:kopka}
-%   J.~Chinju, O.~Matthew,~and~J.~Sahoo. "CNN-LSTM based classification of polo like kinase family of Proteins: An emerging cancer drug target."
-%   \emph{Materials Today: Proceedings}, 2022.
-% \bibitem{IEEEhowto:kopka}
-%   S.~K.~Sønderby~and~O.~Winther, "Protein Secondary Structure Prediction with Long Short Term Memory Networks," 2014,
-%   [Online]. Available: http://arxiv.org/abs/1412.7828.
-% \bibitem{IEEEhowto:kopka}
-%   N.~G.~Nguyen,~V.~A.~Tran,~D.~L.~Ngo et al., “DNA sequence classification by convolutional neural network,” \emph{Journal of Bio-
-%   medical Science and Engineering}, vol. 9, no. 5, pp. 280–286, 2016.
-% \bibitem{IEEEhowto:kopka}
-%   D.~T.~Do~and~N.~Q.~K.~Le, “Using extreme gradient boosting to identify origin of replication in Saccharomyces cerevisiae via
-%   hybrid features,” \emph{Genomics}, vol. 112, no. 3, pp. 2445–2451,2020
-% \bibitem{IEEEhowto:kopka}
-%   X.~Zhang,~B.~Beinke,~B.~Al~Kindhi,~and~M.~Wiering, “Comparing machine learning algorithms with or without feature extraction for DNA classification"
-%   2020, http://arxiv.org/abs/2011.00485
-
-% \end{thebibliography}
 \end{document}

### 2022-05-09 project_discussion_report/main.tex
Fixed typos, improved sentences

@@
     any hyperparameters. All of our results for our different model architectures were generating
-    high accuracy (>98\%) on our single de novo sequence. In the future, we would like to experiment
-    more with tuning hyperparameters in particular if the task becomes harder for different starting
-    de novo sequences and therefore different training datasets. Table~\ref{tab:hyperparameters}
+    high accuracy (98-99\%) on our single de novo sequence. Table~\ref{tab:hyperparameters}
     displays our chosen hyperparameters.
@@
   \subsection{Predictions on De Novo Sequence}
-    To perform the task of deploying our models to predict gaps in our de novo sequence, 
+    To deploy our models to predict gaps in our de novo sequence, 
     we trained two models: one to learn the dependencies in the forward direction and
-    one to learn the dependencies in the reverse direction. This was necessary to deploy 
-    our models to predict gaps in the beginning or end of the target sequence. 
+    one to learn the dependencies in the reverse direction. This was necessary to predict 
+    gaps in the beginning or end of the target sequence. 
     For example, if our target sequence begins with "DIQMSP" and we want to predict "DIQ", then
